Parameter Inference\\with Bifurcation Diagrams

### 2021-05-28 docs/article.tex
submitting to neurips

\begin{document}
\input{notation}
\maketitle
\begin{abstract}
    Estimation of parameters in differential equation models can be achieved by applying learning algorithms to time-series data. However, sometimes it is only possible to measure qualitative changes of a system in response to a controlled condition. In dynamical systems theory, such change points are known as \textit{bifurcations} and lie on a function of the controlled condition called the \textit{bifurcation diagram}. In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The cost function contains a supervised error term that is minimal when the model bifurcations match the specified targets and an unsupervised bifurcation measure which has gradients that push optimisers towards bifurcating parameter regimes. The gradients can be computed without the need to differentiate through the operations of the solver that was used to compute the diagram. We demonstrate parameter inference with minimal models which explore the space of saddle-node and pitchfork diagrams and the genetic toggle switch from synthetic biology. Furthermore, the cost landscape allows us to organise models in terms of topological and geometric equivalence.
\end{abstract}

@@

Inverse problems \cite{Abdulla2009InverseBiology} arise in biology and engineering in settings when the model is not fully known and the desire is to match model behaviour to a given set of observations. This helps systematically guide both model and experimental design. The collected observations, however, are often indirectly proportional to the variables defined in the model. In fluorescence microscopy, a commonly used technique in biological studies, data are reported in arbitrary fluorescence units, allowing the observer to shift and scale data arbitrarily. Models of cellular networks, on the other hand, often seek to describe the dynamics of concentrations of relevant proteins \cite{Tyson2001NetworkPhysiology,Grant2020InterpretationCircuit}. Furthermore, the observed qualitative changes in response to changes in experimental conditions are more robust and reproducible across studies than the quantitative details. For example, several studies are likely to agree that the human immune system activates above a threshold concentration of a pathogen and deactivates at a lower threshold concentration, but may disagree on the exact quantities of the thresholds or the magnitudes of the immune response. Bifurcation theory provides us a framework for studying these qualitative changes in a manner that is independent of quantitative details \cite{Kuznetsov2004TopologicalSystems,Tyson2001NetworkPhysiology}. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring other properties of a system \cite{Stumpf2019ParameterBifurcations}.

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how multiple  parameters in concert control the existence and position of a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made is cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. In the emerging field of Scientific machine learning \cite{Rackauckas2017Differentialequations.jlJulia,Rackauckas2018ASolutions,Rackauckas2020UniversalLearning}, parameters of structured mechanistic models are favoured over flexible models in larger parameter spaces. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

@@

In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equation models that give rise to bifurcations at specified values of a control parameter. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. We apply the strategy of implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients. To compute the diagram we use a predictor-corrector method called deflated continuation \cite{Farrell2016TheDiagrams,Veltz2020BifurcationKit.jl} developed for partial differential equations. In the case of partial differential equations, the computational complexity of calculating even a single bifurcation diagram can be unbounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

@@

\begin{itemize}
    \item An end-to-end differentiable method for locating bifurcations in parameter space and then matching their dependency on a control condition to user-specified locations
    \item Implementation of the method as a Julia package\\  \href{https://github.com/gszep/BifurcationFit.jl}{\texttt{github.com/gszep/BifurcationFit.jl}}
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

@@

For the purposes of introducing this work, we will consider the simplest class of bifurcations known as \textit{co-dimension one} bifurcations not including limit cycles. Therefore $\targets$ should contain conditions for which we hypothesise changes in multi-stable behaviour. Let the equations be
\begin{align}
	\frac{\partial u}{\partial t}=\rates(u,p)
	\qquad\mathrm{where}\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}

In the context of the differential equations, and not considering limit cycles for now, a co-dimension one bifurcation can be defined by a set of conditions on the determinant of the Jacobian $\Det$. The determinant of the Jacobian quantifies the rate at which trajectories in a local patch of state-space $u\in\Reals^N$ converge or diverge. The determinant approaching zero means that the dynamics of the system are slowing down, which is an important indicator for the onset of a transition between qualitative behaviours. Furthermore, the slowing down must necessarily be followed by breakdown of stability; for this to be true it is sufficient to require that the determinant cross zero with a finite slope, meaning that its directional derivative along the diagram $\frac{d}{ds}\Det$ is not zero. The set of predicted values for the control condition $\predictions(\theta)\subset\Reals$ at which bifurcations occur are defined as
\begin{align}
	\predictions(\theta):=\left\{\,
	p\,\,\exists\,\,u:\,\,\rates(u,p)=0,\,\,\Det=0,
	\,\, \frac{d}{ds}\Det\neq0
	\,\right\}
	\label{eq:predictions}
\end{align}

To illustrate how the shapes of bifurcation diagrams compare with the determinant of the Jacobian, we show minimal models for saddle-node (Figure \ref{fig:saddle-node}) and pitchfork bifurcations (Figure \ref{fig:pitchfork}). Indeed the predictions $\predictions(\theta)$ are defined by zero crossings in the determinant with a finite slope. The location of these crossings in general may not match the targets $\targets$.

@@

\clearpage
\section{Proposed Method}
\label{section:method}
\subsection{Semi-supervised Cost Function}

To identify parameter sets that give rise to bifurcation diagrams with specified turning points, we propose a semi-supervised cost function that comprises two terms. The role of the supervised term is simply to reward predicted bifurcations to coincide with the specified target locations. This of course relies on such bifurcations existing. Therefore, the role of the unsupervised term is to encourage an optimiser to move towards parameter regimes that do exhibit bifurcations.

\subsubsection{Supervised term: matched bifurcations to target locations}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. A naive approach might take an average over the norms for each prediction-target pair. However, this can result in multiple predictions matching the same target. Therefore, we choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target (in cases where $|\predictions|\geq|\targets|$):
\begin{equation}
    \error(\theta, \targets) = \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    %\label{eq:error}
\end{equation}
The supervised term is only zero when each target is matched by at least one prediction.

\subsubsection{Unsupervised term: encouraging bifurcations}

\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars.}
    \label{fig:measure}
\end{wrapfigure}

We can see from Figures \ref{fig:minimal-models} and definitions \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(s)$ of zero crossings in the determinant along the bifurcation curve, which we define as 
\begin{equation}
    \measure(s):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{equation}

The bifurcation measure $\measure(s)$ is maximal at bifurcations and has finite gradients in non-bifurcating regimes (Figure \ref{fig:measure}). More specifically, the measure $\measure(s)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero and hence the measure $\measure(s)$ going to zero for each turning point (see Figure \ref{fig:measure}). As the determinant $\Det$ diverges, we approach regimes far away from any closest bifurcation and hence $\measure(s)\rightarrow0$. But we would still like to have non-zero gradients with respect to $\theta$ in this regime, and so the measure is designed to go to zero slowly. Finally, we note that $\measure(s)$ has the desirable property of being invariant to certain classes of transformations on $\rates(u,p)$ that do not change bifurcation locations (details provided in Appendix \ref{appendix:measure}).

While the calculation of the determinant is straightforward, its directional derivative is not. Fortunately, a tangent field representation of the bifurcation curve allows us to compute this derivative implicitly; see Appendix \ref{appendix:tangent-fields} for details.

Equipped with a measure that quantifies the appearance of bifurcations along a bifurcation arc, we now define the total measure for a bifurcation diagram as
\begin{equation}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }.
\end{equation}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations.

The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood. This defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \error(\theta, \targets),
    \label{eq:loss}
\end{align}
where we introduce the hyperparameter $\lambda$ to enable control over the relative contribution of the unsupervised term. The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached.

@@

To make use of gradient-based optimisers to locate desired bifurcation diagrams, we here show how to differentiate the cost function.
First, we note that while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. Therefore, we can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ (within the definition of $\error$) and the total measure $\Psi(\theta)$. Therefore, 
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}

In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

\begin{figure}[ht]
\centering
\setlength\unitlength{1cm}
{\phantomsubcaption\label{fig:two-state-optima:parameters}}
{\phantomsubcaption\label{fig:two-state-optima:models}}
% \includegraphics[width=9cm]{two-state-optima}
\includegraphics[width=9cm]{two-state-optima.png}
\begin{picture}(0,0) \put(-9,6.3){\subref{fig:two-state-optima:parameters}} \end{picture}
\begin{picture}(0,0) \put(-3,6.3){\subref{fig:two-state-optima:models}}
\end{picture}
\caption{Bifurcation inference for the two-state model \eqref{eq:two-state}. \subref{fig:two-state-optima:parameters}. Optimal parameter estimates $\theta^*$ for the targets $\targets=\{4,5\}$ reveal two clusters of qualitatively different regimes: mutual activation ($a_1 < 1$; cluster 1) and mutual inhibition ($a_1 > 1$; cluster 2). \subref{fig:two-state-optima:models}. Example bifurcation diagrams indicate positively and negatively correlated dependencies between the two model states, as a function of the control condition.}
\label{fig:two-state-optima}
\end{figure}

@@

Performing one iteration of the optimisation requires the computation of the gradient of the cost \eqref{eq:gradient-loss}, requiring a computation of the bifurcation diagram with parameter continuation methods, the evaluation of matrix inversions \eqref{eq:gradient-bifurcation}. However instead of evaluating the inversions directly, we solve a system of linear equations, applying the same strategy as implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels}. This leaves us with the computational bottleneck of calculating the determinant of the state space Jacobian, required in the both the bifurcation measure \eqref{eq:measure} and gradient \eqref{eq:gradient-bifurcation}. This calculation scales like $N^3$ where $N$ is the number of state space variables (Figure \ref{fig:scaling}).

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width=0.48\textwidth]{scaling}
\caption{Complexity scaling of calculating the gradient of the cost function. Calculations were performed on an Intel Core i7-6700HQ CPU @ 2.60GHz x 8 without GPU acceleration}
\label{fig:scaling}
\end{wrapfigure}

For this calculation a model \eqref{eq:scaling} was chosen that so that it is extensible both in the number of parameters $M$ and the number of states $N$. In this model only the first state $u_1$ defines the shape of the bifurcation diagram, while the remaining states are merely linearly proportional to the first. The parameters $\mu_n$ contain sums of $\theta_m$ allowing us a flexible choice on the number of parameters while maintaining stable solutions for the bifurcation diagram.

While still tractable on laptop computers for states $N<100$ our implementation currently does not scale well for partial differential equations where a large the number of states $N$ arises from discretisation of the spatial variables. The only reason we need this determinant is because it is an indicator of bifurcations. We can address the computational bottleneck by finding a more computationally efficient way of calculating this indicator. One approach would be to take the product of a finite subset of eigenvalues of the system. Note that any more efficient calculation must still permit back-propagation through it.
\begin{equation}
\begin{cases}
    \,\,\partial_t u_1 = \sin^{2}\!p - (\theta_1\sin^{2}\!p+1)u_1\\
    \,\,\partial_t u_n = u_{n-1} - (\mu_n^2+1) u_n & 2\le n\leq N
\end{cases}
\label{eq:scaling}
\end{equation}
\section{Conclusion \& Broader Impact}
\label{section:conclusion}

We proposed a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. By applying implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} and the generalised Leibniz rule \cite{Flanders1973DifferentiationSign} to the geometry of the implicitly defined steady states \cite{Goldman2005CurvatureSurfaces} it is possible to use automatic differentiation methods to efficiently calculate gradients. We defined a bifurcation measure that that uses the determinant of the state-space Jacobian as an indicator for bifurcating parameter regimes in the unsupervised term of the cost function. The gradients of the cost can be efficiently computed using automatic differentiation methods. The computational bottleneck is the evaluation of the state-space Jacobian determinant which limits the implementation to ordinary differential equations.

>>>>

\begin{document}
\input{notation}
\maketitle
\begin{abstract}
    Estimation of parameters in differential equation models can be achieved by applying learning algorithms to quantitative time-series data. However, sometimes it is only possible to measure qualitative changes of a system in response to a controlled condition. In dynamical systems theory, such change points are known as \textit{bifurcations} and lie on a function of the controlled condition called the \textit{bifurcation diagram}. In this work, we propose a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. The cost function contains a supervised error term that is minimal when the model bifurcations match the specified targets and an unsupervised bifurcation measure which has gradients that push optimisers towards bifurcating parameter regimes. The gradients can be computed without the need to differentiate through the operations of the solver that was used to compute the diagram. We demonstrate parameter inference with minimal models which explore the space of saddle-node and pitchfork diagrams and the genetic toggle switch from synthetic biology. Furthermore, the cost landscape allows us to organise models in terms of topological and geometric equivalence.
\end{abstract}

@@

Inverse problems \cite{Abdulla2009InverseBiology} arise in biology and engineering in settings when the model is not fully known and the desire is to match model behaviour to a given set of observations. This helps systematically guide both model and experimental design. While we would like to understand the quantitative details of a system, often only qualitative changes in response to varying experimental conditions can be robustly measured across independent studies \cite{Tyson2001NetworkPhysiology,Grant2020InterpretationCircuit}. For example, several studies are likely to agree that the human immune system activates above a threshold concentration of a pathogen and deactivates at a lower threshold concentration, but may disagree on the exact quantities of the thresholds or the magnitudes of the immune response. Bifurcation theory provides us a framework for studying these transitions in a manner that is independent of quantitative details \cite{Kuznetsov2004TopologicalSystems}. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring other properties of a system \cite{Stumpf2019ParameterBifurcations}.

Inferring the parameters of a model directly from a bifurcation diagram is difficult because it is not obvious how multiple  parameters in concert control the existence and position of a bifurcation. It could even be impossible for the model to bifurcate in the manner desired. Several approaches exist to place bifurcations to desired locations once a manifold is present \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations} yet typically resort to sampling techniques to search for them in the first place \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}. Progress has been made in cases where model structure and stability conditions are used to refine the search space \cite{Otero-Muras2018Optimization-basedModels,Otero-Muras2014ACurves} yet the resulting objectives are still not explicit in the bifurcation targets and also not differentiable. In the emerging field of scientific machine learning \cite{Rackauckas2017Differentialequations.jlJulia,Rackauckas2018ASolutions,Rackauckas2020UniversalLearning}, parameters of structured mechanistic models are favoured over flexible models in larger parameter spaces. A scalable method for navigating the space of bifurcation diagrams would enable design of differential equations with high-level qualitative constraints. Furthermore one could begin organising models according to qualitatively distinct behaviours.

@@

The problem of inferring differential equation parameters against a user-specified bifurcation diagram decomposes into two parts: searching for bifurcating regimes and matching the locations of bifurcation points to desired values. Matching bifurcation locations is a supervised problem where the data are expressed as bifurcations points \cite{Lu2006InverseSystems,Conrad2006BifurcationClock}. Searching for bifurcations is an unsupervised problem because when bifurcations are not present, there is no distance defined between data and prediction \cite{Chickarmane2005BifurcationTool}. Therefore only properties of the model can be used to start the search. We propose a semi-supervised approach for performing both tasks in an end-to-end fashion. The bifurcation diagram encodes high-level qualitative information defined by state space structures, rather than kinetics. We apply the strategy of implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients. To compute the diagram we use a predictor-corrector method called deflated continuation \cite{Farrell2016TheDiagrams,Veltz2020BifurcationKit.jl} developed for partial differential equations.

@@

\begin{itemize}
    \item An end-to-end differentiable method for locating bifurcations in parameter space and then matching their dependency on a control condition to user-specified locations
    \item Implementation of the method as a Julia package\\  \href{https://github.com/anon/Code.jl}{\texttt{see attached zip file}}
% \href{https://github.com/gszep/BifurcationFit.jl}{\texttt{github.com/gszep/BifurcationFit.jl}}
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

@@

For the purposes of introducing this work, we will consider the simplest class of bifurcations known as \textit{co-dimension one} bifurcations not including limit cycles. Therefore $\targets$ should contain conditions for which we hypothesise changes in multi-stable behaviour. Let the equations be
\begin{align}
	\frac{\partial u}{\partial t}=\rates(u,p)
	\qquad\mathrm{where}\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}
In the context of the differential equations, and not considering limit cycles for now, a co-dimension one bifurcation can be defined by a set of conditions on the determinant of the Jacobian $\Det$. The determinant of the Jacobian quantifies the rate at which trajectories in a local patch of state-space $u\in\Reals^N$ converge or diverge. The determinant approaching zero means that the dynamics of the system are slowing down, which is an important indicator for the onset of a transition between qualitative behaviours. Furthermore, the slowing down must necessarily be followed by breakdown of stability; for this to be true it is sufficient \textit{but not necessary} to require that the determinant cross zero with a finite slope, meaning that its directional derivative along the diagram $\frac{d}{ds}\Det$ is not zero. The set of predicted values for the control condition $\predictions(\theta)\subset\Reals$ at which bifurcations occur are defined as
\begin{align}
	\predictions(\theta):=\left\{\,
	p\,\,\exists\,\,u:\,\,\rates(u,p)=0,\,\,\Det=0,
	\,\, \frac{d}{ds}\Det\neq0
	\,\right\}
	\label{eq:predictions}
\end{align}
The most common bifurcations between steady states, not including limit cycles, are saddle-nodes and pitchforks \cite{Haragus2011LocalSystems}. The relationships between our conditions for bifurcation \eqref{eq:predictions} and the conditions for saddle-node are detailed in Appendix \ref{appendix:conditions}. Saddle-node bifurcations, which often appear in pairs (Figure \ref{fig:saddle-node}) are defined by stable and unstable fixed points meeting and disappearing. Pitchfork bifurcations are special cases of the saddle-node where a single steady state splits into two stable and one unstable steady state (Figure \ref{fig:pitchfork}). To illustrate these bifurcations, we define minimal models (Figure \ref{fig:minimal-models}) that span the space of saddle-node and pitchforks, where indeed zero crossings in the determinant with a finite slope define the set of prediction $\predictions(\theta)$. The location of these crossings in general may not match the targets $\targets$.

@@

\section{Proposed Method}
\label{section:method}
\subsection{Semi-supervised Cost Function}

To identify parameter sets that give rise to bifurcation diagrams with specified bifurcation points, we propose a semi-supervised cost function that comprises two terms. The role of the supervised term is simply to reward predicted bifurcations to coincide with the specified target locations. This of course relies on such bifurcations existing. The role of the unsupervised term is to encourage an optimiser to move towards parameter regimes that do exhibit bifurcations.

\subsubsection{Supervised term: matching bifurcations to target locations}

In order for predicted bifurcations $p(\theta)\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. A naive approach might take an average over the norms for all prediction-target pairs. However this gives rise to unwanted cross-terms and the possibility of multiple predictions matching the same target without any penalty for unmatched targets. Therefore, we choose a geometric mean over the predictions and an arithmetic mean over targets:
\begin{equation}
    \error(\theta, \targets) = \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions(\theta)}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:error}
\end{equation}
The supervised term is only zero when each target is matched by at least one prediction and allows for cases where the number of predictions is greater than or equal to the number of targets $|\predictions| \geq |\targets|$. An alternative approach, which undesirably introduces more hyper-parameters, would be to let each prediction $\predictions(\theta)$ represent the centroid of a mixture distribution and use expectation-maximisation to match the centroids to targets $\targets$.

\subsubsection{Unsupervised term: encouraging bifurcations}
\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(s)$ and determinant $\Det$ along the arclength $s$ of two different bifurcation curves demonstrating how maximising the measure along the curve maintains the existing bifurcation marked by a circle, while encouraging new bifurcations marked by stars.}
    \label{fig:measure}
\end{wrapfigure}

We can see from Figure \ref{fig:minimal-models} and definition \eqref{eq:predictions} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant crosses zero $\Det=0$ with a finite slope $\frac{d}{ds}\Det\neq 0$. Using these quantities we can define a positive semi-definite measure $\measure(s)$ of zero crossings in the determinant along a curve parameterised by $s$ which we define as 
\begin{equation}
    \measure(s):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{equation}

The bifurcation measure $\measure(s)$ is maximal at bifurcations and has finite gradients in non-bifurcating regimes (Figure \ref{fig:measure}). More specifically, the measure $\measure(s)$ is one at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in the directional derivative $\frac{d}{ds}\Det$ going to zero. Hence the measure $\measure(s)$ goes to zero for each turning point (see Figure \ref{fig:measure}).

On the other hand, as the determinant $\Det$ diverges, we approach regimes far away from any bifurcations and hence $\measure(s)\rightarrow0$. Since we would still like to have non-zero gradients with respect to $\theta$ in these regimes we designed the measure to go to zero sufficiently slowly.
\clearpage

% Finally, we note that $\measure(s)$ has the desirable property of being invariant to certain classes of transformations on $\rates(u,p)$ that do not change bifurcation locations (details provided in Appendix \ref{appendix:measure}).

While the calculation of the determinant is straightforward, its directional derivative requires a tangent vector to the bifurcation curve. Fortunately the tangent vector $\tangent(s)$ at the solution $u(s),p(s)$ anywhere along the curve $s$ can be calculated as the nullspace of the rectangular $N\times(N+1)$ Jacobian
\begin{equation}
    \left.
    \frac{\partial\rates}{\partial(u,p)}
    \right|_{\rates(u(s),p(s))=0}
    \cdot\tangent(s)=0
    \label{eq:tangent}
\end{equation}
This equation guarantees that the tangent vector $\tangent(s)$ is orthogonal to all hyper-planes defined by the components of $\rates$. In this setting the dimension of the nullspace is always known, and therefore can reliably be calculated using QR factorisation methods \cite{Drmac2008OnStudy}.

Equipped with a measure that quantifies the appearance of bifurcations along a bifurcation arc we can define the total measure for a bifurcation diagram as
\begin{equation}
    \Psi(\theta):=\frac{
        \int_{\rates(u,p)=0}\!
        \measure(s)\,\mathrm{d}s
    }{
        \int_{\rates(u,p)=0}\!\
        \mathrm{d}s
    }.
    \label{eq:total-measure}
\end{equation}
Here we denote $\int_{\rates(u,p)=0}\mathrm{d}s$ as the sum of the line integrals in $(u,p)\in\Reals^{N+1}$ defined by the level set $\rates(u,p)=0$ with $s$ being an arbitrary parametrisation of the curves. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the controlled condition region $p$ is densely packed with bifurcations. The total measure $\Psi(\theta)$ is added to the supervised term as if it were a likelihood. This defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\log\Psi(\theta)+
    \error(\theta, \targets),
    \label{eq:loss}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached.

@@

To make use of gradient-based optimisers to locate desired bifurcation diagrams, we show here how to differentiate the cost function. First, we note that while individual bifurcations $p(\theta)$ depend smoothly on $\theta$, the total number of predictions $|\predictions|$ does not have gradient contributions with respect to $\theta$. Therefore, we can safely drop the dependency in the prediction counter and now proceed in taking gradients with respect to $\theta$ knowing that the only dependencies we need to track are for individual bifurcations $p(\theta)$ within the definition the supervised term \eqref{eq:error} and the total measure \eqref{eq:total-measure}. Therefore, 
\begin{align}
    \frac{\partial\loss}{\partial\theta}=
    \big(|\predictions|-|\targets|\big)\,\lambda\,
    \frac{\partial \Psi}{\partial\theta}\Psi(\theta)^{-1}+
    \frac{1}{|\targets||\predictions|}\sum_{p'}
    \prod_{p(\theta)}|p(\theta)-p'|^{\frac{1}{|\predictions|}}
    \sum_{p(\theta)}\frac{\partial p}{\partial\theta}\left(p(\theta)-p'\right)^{-1}
    \label{eq:gradient-loss}
\end{align}
In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to calculate the gradient $\frac{\partial\loss}{\partial\theta}$ without having to differentiate through the operations of the solver that finds the bifurcation diagram $\rates(u,p)=0$ and the bifurcation locations $p(\theta)$. To calculate the gradient of the measure $\frac{\partial \Psi}{\partial\theta}$ we need to differentiate line integrals that depend on $\theta$. Fortunately this can be done by the application of the generalised Leibniz integral rule, details of which can be found in Appendix \ref{appendix:leibniz-rule}.

@@

\begin{figure}[ht]
\centering
\setlength\unitlength{1cm}
{\phantomsubcaption\label{fig:two-state-optima:parameters}}
{\phantomsubcaption\label{fig:two-state-optima:models}}
\includegraphics[width=9cm]{two-state-optima}
\begin{picture}(0,0) \put(-9,6.3){\subref{fig:two-state-optima:parameters}} \end{picture}
\begin{picture}(0,0) \put(-3,6.3){\subref{fig:two-state-optima:models}}
\end{picture}
\caption{Bifurcation inference for the two-state model \eqref{eq:two-state}. \subref{fig:two-state-optima:parameters}. Optimal parameter estimates $\theta^*$ for the targets $\targets=\{4,5\}$ reveal two clusters of qualitatively different regimes: mutual activation ($a_1 < 1$; cluster 1) and mutual inhibition ($a_1 > 1$; cluster 2). \subref{fig:two-state-optima:models}. Example bifurcation diagrams indicate positively and negatively correlated dependencies between the two model states, as a function of the control condition.}
\label{fig:two-state-optima}
\end{figure}

@@

Performing one iteration of the optimisation requires the computation of the gradient of the cost \eqref{eq:gradient-loss}, requiring a computation of the bifurcation diagram with parameter continuation methods, which includes the evaluation of matrix inversions \eqref{eq:gradient-bifurcation}. Instead of evaluating the inversions directly, we solve a system of linear equations, applying the same strategy as implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels}. This leaves us with the computational bottleneck of calculating the determinant of the state space Jacobian, required in both the bifurcation measure \eqref{eq:measure} and gradient \eqref{eq:gradient-bifurcation}. This calculation scales like $N^2$ where $N$ is the number of state space variables (Figure \ref{fig:scaling}).

\begin{figure}[ht]%{r}{0.5\textwidth}
\centering
\includegraphics[width=0.48\textwidth]{scaling}
\caption{Complexity scaling of calculating the gradient of the cost function. Calculations were performed on an Intel Core i7-6700HQ CPU @ 2.60GHz x 8 without GPU acceleration}
\label{fig:scaling}
\end{figure}

For this calculation a model \eqref{eq:scaling} was chosen that so that it is extensible both in the number of parameters $M$ and the number of states $N$. 
\begin{equation}
\begin{cases}
    \,\,\partial_t u_1 = \sin^{2}\!p - (\theta_1\sin^{2}\!p+1)u_1\\
    \,\,\partial_t u_n = u_{n-1} - (\mu_n^2+1) u_n & 2\le n\leq N
\end{cases}
\label{eq:scaling}
\end{equation}
In this model only the first state $u_1$ defines the shape of the bifurcation diagram, while the remaining states are merely linearly proportional to the first. The parameters $\mu_n$ contain sums of $\theta_m$ allowing us a flexible choice on the number of parameters while maintaining stable solutions for the bifurcation diagram.

While still tractable on laptop computers for states $N<100$ our implementation currently does not scale well for partial differential equations where a large the number of states $N$ arises from discretisation of the spatial variables. The only reason we need this determinant is because it is an indicator of bifurcations. We can address the computational bottleneck by finding a more computationally efficient way of calculating this indicator. One approach would be to take the product of a finite subset of eigenvalues of the system. Note that any more efficient calculation must still permit back-propagation through it.

\section{Conclusion \& Broader Impact}
\label{section:conclusion}

We proposed a gradient-based semi-supervised approach for inferring the parameters of differential equations that produce a user-specified bifurcation diagram. By applying implicit layers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} and the generalised Leibniz rule \cite{Flanders1973DifferentiationSign} to the geometry of the implicitly defined steady states \cite{Goldman2005CurvatureSurfaces} it is possible to use automatic differentiation methods to efficiently calculate gradients. We defined a bifurcation measure that uses the determinant of the state-space Jacobian as an indicator for bifurcating parameter regimes in the unsupervised term of the cost function. The gradients of the cost can be efficiently computed using automatic differentiation methods. The computational bottleneck is the evaluation of the state-space Jacobian determinant which limits the implementation to ordinary differential equations.
